---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Unzip the activity.zip file and load data using read.csv.
```{r}
data <- read.csv(unz("activity.zip", "activity.csv"), colClasses=c("integer", "Date", "integer"))
data$interval <- 60*floor((data$interval+1)/100) + (data$interval %% 100)

str(data)
summary(data)
```

## What is mean total number of steps taken per day?
Calculate the mean total steps using the *tapply* function
```{r}
total_steps <- tapply(data$steps, data$date, sum, na.rm=T)
mean_steps <- mean(total_steps)
mean_steps
```

Calculate the median 
```{r}
median_steps <- median(total_steps)
median_steps
```

Generate the histogram and show the mean and median in it.

```{r}
hist(total_steps, breaks=10, xlab="No. of steps/day")
abline(v=mean_steps, col="green", lwd=2)
abline(v=median_steps, col="blue", lwd=2)
legend(x="topright", legend=c("mean", "median"), col=c("green", "blue"), bty="n", lwd=3)
```


## What is the average daily activity pattern?
Plot the activity pattern as a time series.
```{r}
avg_steps <- tapply(data$steps, data$interval, mean, na.rm=T)

hours <- as.numeric(names(avg_steps))/60
plot(hours, avg_steps, type="l", axes=F,
     xlab="time of day (h)", ylab="average number of steps in 5-min interval",
     main="Daily activity pattern")
axis(2)
axis(1, at=0:6*4, labels=paste(0:6*4,":00", sep=""))
```

Find the 5 minute interval that has the maximum number of steps.
```{r}
max_activity_num <- which(avg_steps == max(avg_steps))
max_activity_int <- data$interval[max_activity_num]
sprintf("%02d:%02d", floor(max_activity_int/60), max_activity_int%%60)
```
## Imputing missing values
Check the number of NAs
```{r}
sum(is.na(data))
```
Create a new data frame called impute without the NAs. Using the average number of steps in that 5 minute interval
```{r}
impute <- transform(data, steps=ifelse(is.na(steps), avg_steps, steps))
summary(impute)
```

Find out the total steps using the impute dataset
```{r}
impute_totalsteps <- tapply(impute$steps, impute$date, sum, na.rm=T)
```
Calculate the mean using impute
```{r}
impute_meansteps <- mean(impute_totalsteps)
impute_meansteps
```

Calculate the median as well
```{r}
impute_mediansteps <- median(impute_totalsteps)
impute_mediansteps
```

Create the histogram using impute dataset.
```{r}
hist(impute_totalsteps, breaks=11, 
     xlab="number of steps/day", 
     sub="(missing values imputed)")
abline(v=impute_meansteps, col="green", lwd=2)
abline(v=impute_mediansteps, col="blue", lwd=2, lty=2)
legend(x="topright", legend=c("mean","median"), col=c("green","blue"), bty="n", lwd=2)
```

Check if and by how much the total number of steps have increased because of imputation.
```{r}
sum(data$steps, na.rm=TRUE)
sum(impute$steps)
```
Because of the imputation, the total number of steps has increased.

## Are there differences in activity patterns between weekdays and weekends?
Classify the data as weekend and weekday. Then aggregate by five minute interval and then plot.

```{r}
week <- factor(weekdays(impute$date) %in% c("Saturday","Sunday"), 
               labels=c("weekday","weekend"), ordered=FALSE)

impsteps <- aggregate(impute$steps, by=list(interval=impute$interval, weekday=week), mean)

library(ggplot2)
g <- ggplot(impsteps, aes(interval/60, x))
g + geom_line() + facet_grid(weekday ~ .) +
    scale_x_continuous(breaks=0:6*4, labels=paste(0:6*4,":00", sep="")) +
    theme_bw() +
    labs(y="average number of steps in 5-min interval") +
    labs(x="time of day (h)") +
    labs(title="Daily activity pattern")
```